---
title: 'Reproducible Research: Course Project I'
author: "Talha Mahmood Chaudhry"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Activity Data Analysis


This Markdown file contains code to run analysis on **activity data**.
Make sure that you have the activity data file in your working directory,
for running R code.

### Loading and preprocessing of data

```{r activity}
unzip(zipfile="activity.zip")
activity <- read.csv("./activity.csv")
library(ggplot2)
```

### Histogram of total number of steps taken per day

```{r histogram}
totalsteps <- with(activity, tapply(steps, date, sum, na.rm = TRUE))
qplot(totalsteps, binwidth = 1000, fill = I("blue"),
      xlab = "Total number of steps taken each day", ylab = "Number of days")
```

### Mean and Median of total number of steps per day

```{r mean median}
st.mean <- mean(totalsteps, na.rm = TRUE)
st.median <- median(totalsteps, na.rm = TRUE)
options(digits = 2)
```
The mean is `r st.mean` steps and median is `r st.median` steps.

### Time series plot of average steps taken

```{r time series plot}
ave.steps <- aggregate(x = list(steps = activity$steps),
                       by = list(interval = activity$interval),
                       FUN = mean, na.rm = TRUE)
ggplot(data = ave.steps, aes(interval,steps)) +
  geom_line(color = "blue") +
  labs(x = "5 minute Intervals",
       y = "Average number of steps across all days")
```

#### 5-minute interval with most number of steps on average
```{r interval steps}
max.st <- ave.steps[which.max(ave.steps$steps), ][1]
```
The 5-minute interval is `r max.st`.

### Impute missing values

#### Total number of missing values
```{r NAs}
miss.value <- is.na(activity$steps)
table(miss.value)
```
#### Impute those missing values
Take the average steps taken across 5 minute intervals, and replace
with missing values of steps for 5 minute intervals
```{r impute values}
imp_act <- activity
imp.values <- function(steps, interval) {
  values <- NA
  if (!is.na(steps))
    values <- c(steps)
  else
    values <- (ave.steps[ave.steps$interval==interval, "steps"])
  return(values)
  }
imp_act$steps <- mapply(imp.values, imp_act$steps, imp_act$interval)
```

### Histogram of the total number of steps taken per day with imputed data

```{r imputed histogram}
totalsteps <- with(imp_act, tapply(steps, date, sum, na.rm = TRUE))
qplot(totalsteps, binwidth = 1000, fill = I("red"),
      xlab = "Total number of steps taken each day", ylab = "Number of days")
```

#### Mean and Median of total steps from imputed data
```{r imp.mean.median}
st.mean <- mean(totalsteps, na.rm = FALSE)
st.median <- median(totalsteps, na.rm = FALSE)
options(scipen = 999, digits = 2)
```
The new mean is `r st.mean`, and median is `r st.median`. These values are 
higher than previous values as expected. With the replacement of missing(NA)
values both mean and median have increased.

### Activity patterns between weekdays and weekends

From the date variable the days of the week are inferred. Then for days,
Monday to Friday, the value of new column Days in the imputed data is
assigned *Weekday*. Saturday and Sunday are assigned *Weekend*. Then
a two-panel time series graph is plotted to contrast daily activity patterns
on average across 5-minute intervals

```{r week.comparison}
imp_act$date <- as.Date(imp_act$date)
imp_act$day <- weekdays(imp_act$date)

which.day <- function(x) {
  weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
  if (x %in% weekday)
    paste("Weekday")
  else
    paste("Weekend")
}

imp_act$day <- sapply(imp_act$day, which.day)

ave.steps <- aggregate(steps ~ interval + day, imp_act, mean)

ggplot(ave.steps, aes(interval, steps)) + geom_line(color = "red") +
  facet_grid(day ~ .) +
  labs(x = "5 minute intervals", y = "Number of steps")
```

